ODGI: scalable tools for pangenome graphs

### 2021-06-01 manuscript/main.tex
Fixed topics

    \odgi\ is entirely focused on "first-order" operations on the graph
    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
    graphs from raw sequences, map reads, or call variants. It
    supports interrogation of the graph for basic dimensional
    properties, positional and path-related queries, and visualization
    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
    operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the
    tools are designed to be applied together, piping the output of
    one tool into the next, thereby preventing the creation of
    intermediate files, and reducing the number of IO operations.

@@

    \topic{odgi build, view \& validate} pangenome graphs are
    represented by a subset of the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, such as
    validating that paths respect the graph's topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualization
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome and is
    capable of handling full length human chromosomes
    (Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
    simplifies the graph by applying several algorithms, including the
    novel `Path-Guided Stochastic Gradient Descent' algorithm.
    \cmd{odgi layout} generates a file with the X-Y coordinates for
    all nodes and \cmd{odgi draw} draws the actual layout of the graph
    (Fig.~\ref{fig:1}b).

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into genome complexity in other
    ways. E.g., \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarizes path
    information into bins of a specified size, thereby enabling a
    summarized view of gigabase scale graphs. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow detection of complex
    genomic regions.

    \topic{odgi break, groom, chop \& unchop} \odgi\ pangenome graphs are dynamic structures in ODGI.
    \cmd{odgi break} removes cycles in the graph, reducing its topology complexity. \cmd{odgi groom}
    removes spurious inverting links by exploring the graph from the orientation supported by most paths. \cmd{odgi chop}
    divides long nodes into shorter ones at a maximum requested size, simplifying downstream analysis.
    \cmd{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the
    embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

    \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
    components. \cmd{odgi explode} separates them in different files, while \cmd{odgi squeeze} merges
    multiple graphs into the same file, preventing node ID collisions. \cmd{odgi extract} extracts regions of the graph
    as defined by query criteria, thereby reducing the resources to work only with the subgraph.

    \topic{odgi position} embedded paths provide stable coordinates regardless of the graph in which they are.
    \cmd{odgi position} finds, translates, and liftovers graph and path positions between different graphs by
    exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).

>>>>

    \odgi\ is entirely focused on "first-order" operations on the graph
    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
    graphs from raw sequences, map reads, or call variants. It
    supports interrogation of the graph for basic dimensional
    properties, positional and path-related queries, and visualisation
    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
    operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the
    tools are designed to be applied together, piping the output of
    one tool into the next, thereby preventing the creation of
    intermediate files, and reducing the number of IO operations.

@@

    \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, such as
    validating that paths respect the graph's topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome and is
    capable of handling full length human chromosomes
    (Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
    simplifies the graph by applying several algorithms, including the
    novel `Path-Guided Stochastic Gradient Descent' algorithm.
    \cmd{odgi layout} generates a file with the X-Y coordinates for
    all nodes and \cmd{odgi draw} draws a layout of the graph
    (Fig.~\ref{fig:1}b).

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into genome complexity in other
    ways. E.g., \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow detection of complex
    genomic regions.

    \topic{odgi break, groom, chop \& unchop} pangenome graphs can be
    reorganised.  \cmd{odgi break} removes circular structures in the
    graph, thereby reducing complexity. \cmd{odgi groom} removes
    spurious inverted links by aligning the graph from the orientation
    that is supported by most paths. \cmd{odgi chop} divides long
    nodes into shorter ones at a maximum requested size, thereby
    simplifying downstream analysis.  \cmd{odgi unchop} joins nodes
    and embedded sequences that do not change the graph topology,
    thereby obtaining a more compact representation of the graph.

    \topic{odgi explode, squeeze \& extract} pangenome are constructed
    as a large graph. \cmd{odgi explode} separates units, such as
    chromosomes, into different files.  \cmd{odgi squeeze} merges
    multiple graphs into the same file whilst preventing node ID
    collisions. \cmd{odgi extract} extracts regions of the graph as
    defined by certain criteria, allowing downstream processing of
    smaller subgraphs.

    \topic{odgi position} pangenome graphs are flexible when it comes
    to coordinate systems. \cmd{odgi position} can use the coordinate
    system from a contained reference genome --- a dynamic liftover
    --- to display coordiates and other localised features, as is shown in
    Fig.~\ref{fig:1b}.
